\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{parskip}

\geometry{margin=2.5cm}

\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\Lcal}{\mathcal{L}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\R}{\mathbb{R}}

\title{%
  Position-Resolved Likelihood Inference\\
  for Atom Gradiometer Image Data%
}
\author{}
\date{}

\begin{document}
\maketitle
\tableofcontents
\newpage

%%============================================================
\section{Setup and Notation}
%%============================================================

\subsection{Gradiometer Architecture}

We consider an atom gradiometer composed of two atom interferometers (AIs)
operated simultaneously, one with atoms sourced at height $z = 0$\,m and one at
$z = 100$\,m, indexed hereafter by $z \in \mathcal{Z} = \{0,\,100\}$.
Each AI independently measures the accumulated phase along its atom trajectories.
Common-mode laser-phase noise cancels in the differential signal, while a
spatially-dependent differential phase (e.g.\ from a gravitational wave or
dark-matter field) is retained.

A run consists of $N_\mathrm{shots}$ shots indexed $i = 1,\ldots,N_\mathrm{shots}$,
acquired at times $t_i$.  Each shot produces two 2D fluorescence images per AI:
one for ground-state atoms ($s = g$, bright port) and one for excited-state atoms
($s = e$, dark port).  Images share a common pixel grid with bin centres
$\x_b = (x_b, y_b)$, bin area $\Delta A = \Delta x\,\Delta y$, and
$N_b$ bins total.

\subsection{Parameters}

\paragraph{Science signal.}
The target quantity is the differential phase signal injected into AI $z=100$:
\begin{equation}
  \delta\phi_i(\beta)
  = A_s \sin(2\pi f\,t_i) + A_c \cos(2\pi f\,t_i),
  \qquad \beta = (A_s,\,A_c),
  \label{eq:signal}
\end{equation}
where the signal frequency $f$ is assumed known a priori, leaving two free
parameters.  The signal amplitude and phase are
$A = \sqrt{A_s^2 + A_c^2}$ and $\varphi = \arctan2(A_c, A_s)$.

\paragraph{Common laser phase.}
The common laser phase $\phi_i \sim \mathrm{Uniform}(0, 2\pi)$ is shared by
both AIs within each shot, drawn independently across shots, and unobserved.
It is a latent variable; how we handle it in the likelihood is discussed in
Section~\ref{sec:nuisances}.

\paragraph{Effective AI phase.}
The total phase entering the transition probability for AI $z$ in shot $i$ is
\begin{equation}
  \Phi_{iz}(\phi_i,\beta)
  = \phi_i + \psi_z\!\bigl(\delta\phi_i(\beta)\bigr),
  \qquad
  \psi_0 = 0,\quad \psi_{100} = \delta\phi_i(\beta).
  \label{eq:eff_phase}
\end{equation}

\paragraph{Cloud nuisance parameters.}
The initial atomic cloud for AI $z$ in shot $i$ is characterised by a
Gaussian phase-space distribution with parameters
\begin{equation}
  \eta_{iz}
  =
  \bigl(
    \mu_{x_0},\,\mu_{y_0},\,\mu_{v_x},\,\mu_{v_y},\,
    \sigma_x,\,\sigma_y,\,\sigma_{v_x},\,\sigma_{v_y}
  \bigr)_{iz},
\end{equation}
the transverse centre-of-mass (COM) positions, COM velocities, and Gaussian
widths.  These vary shot-to-shot and across AIs.  Like $\phi_i$, they are
unobserved and must be handled carefully in the likelihood; see
Section~\ref{sec:nuisances}.

%%============================================================
\section{Semiclassical Phase-Space Map (PSMAP)}
%%============================================================

\subsection{PSMAP Surrogate}

The semiclassical simulation (ais++) integrates the interferometer sequence for a
regular 4D grid of initial transverse phase-space coordinates
$\q_0 = (x_0, y_0, v_{x0}, v_{y0})$ at fixed source height and longitudinal
velocity.  For each grid point and each output port $p$ it stores:
(i) wavepacket path amplitudes $a_{0,p}(\q_0)$ and $a_{1,p}(\q_0)$;
(ii) accumulated phase difference $\Delta\phi_p(\q_0)$;
(iii) interference flag $\iota_p \in \{0,1\}$; and
(iv) output state label $s_p \in \{g,e\}$.
At runtime, values are obtained by quadrilinear interpolation on the
$21^4$ grid (\texttt{PSMAPSurrogate}).
A separate PSMAP is computed for each AI.

\subsection{Transition Probabilities}

The probability that an atom with initial condition $\q_0$ is detected at port
$p$ given total AI phase $\Phi$ is
\begin{equation}
  P_p(\q_0;\,\Phi)
  = a_{0,p}^2 + a_{1,p}^2
    + \iota_p \cdot 2\,a_{0,p}\,a_{1,p}\,
      \cos\!\bigl(\Delta\phi_p(\q_0) + \Phi\bigr).
  \label{eq:port_prob}
\end{equation}
The per-state probabilities are obtained by summing over ports with matching state
label:
\begin{equation}
  p_s(\q_0;\,\Phi)
  = \sum_{p:\,s_p = s} P_p(\q_0;\,\Phi),
  \qquad s \in \{g, e\}.
  \label{eq:state_prob}
\end{equation}
Because the simulation retains only the two dominant interferometry paths,
\begin{equation}
  p_g(\q_0;\,\Phi) + p_e(\q_0;\,\Phi) \leq 1;
  \label{eq:pruning}
\end{equation}
atoms not accounted for are lost to pruned paths.  In particular,
$p_e \neq 1 - p_g$ in general; both are computed independently.

\subsection{Ballistic Trajectory Map}

Under free-flight propagation over detection time $T_\mathrm{det}$,
the final transverse position is
\begin{equation}
  \x^f(\q_0)
  =
  \begin{pmatrix}
    x_0 + T_\mathrm{det}\,v_{x0} \\
    y_0 + T_\mathrm{det}\,v_{y0}
  \end{pmatrix}.
  \label{eq:ballistic}
\end{equation}
Both output states follow the same trajectory (no state-dependent force before
imaging).  With default parameters ($T_\mathrm{det} = 3.8$\,s,
$\sigma_x = 100\,\mu$m, $\sigma_{v_x} = 309\,\mu$m/s), the cloud expands to
$\sigma_{x^f} \approx 1.2$\,mm at detection.

%%============================================================
\section{Per-Pixel Statistics via QMC}
\label{sec:qmc}
%%============================================================

\subsection{ACS Decomposition}

Expanding the cosine in~\eqref{eq:port_prob}, the phase dependence of $p_s$ is
entirely captured by three $\Phi$-independent scalars per pixel $b$, AI $z$, state $s$:
\begin{align}
  A_{zsb}(\eta_{iz})
  &= \E_{\q_0 \sim \text{cloud},\, \x^f\in b}
     \!\bigl[a_{0}^2(\q_0) + a_{1}^2(\q_0)\bigr],
  \label{eq:Ab} \\[4pt]
  C_{zsb}^c(\eta_{iz})
  &= \E_{\q_0 \sim \text{cloud},\, \x^f\in b}
     \!\bigl[2\,\iota\,a_{0}(\q_0)\,a_{1}(\q_0)\cos\Delta\phi(\q_0)\bigr],
  \label{eq:Cc} \\[4pt]
  C_{zsb}^s(\eta_{iz})
  &= \E_{\q_0 \sim \text{cloud},\, \x^f\in b}
     \!\bigl[2\,\iota\,a_{0}(\q_0)\,a_{1}(\q_0)\sin\Delta\phi(\q_0)\bigr],
  \label{eq:Cs}
\end{align}
where the expectation is over initial conditions $\q_0$ drawn from the Gaussian
cloud $\N(\eta_{iz})$ whose final positions $\x^f(\q_0)$ fall within pixel $b$.
The expected pixel count is then
\begin{equation}
  \boxed{
  \lambda_{izsb}(\Phi)
  \;=\;
  L_{iz}\;
  \bigl[A_{zsb} + C_{zsb}^c\cos\Phi + C_{zsb}^s\sin\Phi\bigr],
  }
  \label{eq:lambda_analytic}
\end{equation}
where $\Phi = \Phi_{iz}(\phi_i,\beta)$ from~\eqref{eq:eff_phase} and $L_{iz}$
is an atom-number scale factor (Section~\ref{sec:atom_number}).

\subsection{QMC Evaluation (\texttt{SurrogatePixelACS})}

The conditional expectations~\eqref{eq:Ab}--\eqref{eq:Cs} are evaluated by
quasi-Monte Carlo (QMC) quadrature with $n_\mathrm{quad}$ Sobol points.

\begin{enumerate}
  \item Draw $n_\mathrm{quad}$ quasi-random standard-normal deviates
        $\mathbf{z}_j \in \R^4$, $j=1,\ldots,n_\mathrm{quad}$, using a
        scrambled Sobol sequence.  These are fixed at construction time and
        reused across all $\theta$ evaluations.

  \item Transform to cloud Gaussian: $\q_{0,j} = \mu(\eta) + \mathrm{diag}(\sigma(\eta))\,\mathbf{z}_j$.

  \item Compute detection positions $\x^f_j$ via~\eqref{eq:ballistic} and
        assign each sample to pixel $b_j$ using pixel edge arrays.
        Samples outside the image boundary are discarded.

  \item Evaluate the PSMAP at all $\q_{0,j}$ simultaneously via vectorised
        quadrilinear interpolation on the \texttt{PSMAPSurrogate} grid.
        This yields amplitudes $(a_{0,j}, a_{1,j})$ and phase residuals
        $\Delta\phi_j$ for all ports, in one GPU kernel with no Python loops.

  \item Extract $(A,C^c,C^s)$ per point using a \textbf{3-phase trick}:
        evaluate $p_s(\q_{0,j};\,\Phi)$ at $\Phi \in \{0, \pi/2, \pi\}$,
        and solve for $A$, $C^c$, $C^s$ by
        \begin{equation}
          A = \tfrac{1}{2}[p(\Phi=0) + p(\Phi=\pi)],\quad
          C^c = \tfrac{1}{2}[p(\Phi=0) - p(\Phi=\pi)],\quad
          C^s = p(\Phi=\pi/2) - A.
        \end{equation}

  \item Accumulate per-pixel ACS by equal-weight bincount:
        \begin{equation}
          A_{zsb} = \frac{1}{n_\mathrm{in}}
                    \sum_{j:\,b_j = b} A_{j},
        \end{equation}
        and analogously for $C^c_{zsb}$, $C^s_{zsb}$, where
        $n_\mathrm{in}$ is the number of samples inside the detector.
        Both ground and excited state statistics are accumulated simultaneously.
\end{enumerate}

The default $n_\mathrm{quad} = 20{,}000$ Sobol points (rounded up to the next
power of 2) gives ACS tensors with sub-percent statistical noise in occupied pixels,
at a cost of one vectorised PSMAP evaluation per call.

%%============================================================
\section{Generative Model and Likelihood}
%%============================================================

\subsection{Poisson Observation Model}

Conditional on $\phi_i$ and $\eta_{iz}$, pixel counts are independent Poisson:
\begin{equation}
  n_{izsb} \;\sim\; \mathrm{Poisson}\!\bigl(\lambda_{izsb}(\Phi_{iz})\bigr),
  \qquad s \in \{g,e\},\; b = 1,\ldots,N_b.
  \label{eq:poisson}
\end{equation}
The per-shot conditional log-likelihood (for both AIs jointly) is
\begin{equation}
  \log\Lcal_i(\beta,\phi_i,\{\eta_{iz}\})
  = \sum_{z,s,b}
    \Bigl[
      n_{izsb} \log \lambda_{izsb}(\Phi_{iz})
      - \lambda_{izsb}(\Phi_{iz})
    \Bigr],
  \label{eq:cond_ll}
\end{equation}
where $\log(n!)$ constants are dropped as parameter-free.

\subsection{Atom Number Scale Factor}
\label{sec:atom_number}

Rather than tracking the launched atom count $\Lambda_{iz}$ explicitly, we
absorb it into a per-shot, per-AI scale factor $L_{iz}$ defined by the
requirement that predicted total counts match observed total counts at the
profiled cloud parameters $\hat\eta_{iz}$:
\begin{equation}
  L_{iz}
  = \frac{\displaystyle\sum_b \bigl(n_{izgb} + n_{izeb}\bigr)}
         {\displaystyle\sum_b \bigl(A_{zsb}(\hat\eta_{iz})
                              + A_{zsb}^{(e)}(\hat\eta_{iz})\bigr)},
  \label{eq:Lfix}
\end{equation}
where the denominator sums the $\phi$-independent terms over all pixels for
both states.  This normalization is exact whenever the interference terms
$C^c$ and $C^s$ integrate to zero over all pixels (complementarity), which
holds to high accuracy in practice.  $L_{iz}$ is computed once at
$\hat\eta_{iz}$ and treated as fixed.

%%============================================================
\section{Handling Unobserved Parameters: Profiling and Marginalisation}
\label{sec:nuisances}
%%============================================================

The full likelihood~\eqref{eq:cond_ll} involves two kinds of unobserved parameters
per shot: the common laser phase $\phi_i$, and the cloud parameters $\eta_{iz}$.
Both must be eliminated before optimising over $\beta$.

\subsection{Marginalise $\phi_i$}

$\phi_i \sim \mathrm{Uniform}(0, 2\pi)$ is completely unknown.  Profiling
(jointly optimising $\phi_i$ with $\beta$) would allow the per-shot phase to
absorb the signal entirely, biasing $\hat\beta$ toward zero.  We therefore
marginalise numerically using a uniform phase grid
$\phi_k = 2\pi k / K$, $k = 0,\ldots,K-1$:
\begin{equation}
  \log\Lcal_i^\mathrm{marg}(\beta,\{\hat\eta_{iz}\})
  = \operatorname{logsumexp}_{k=0}^{K-1}
    \!\bigl[\log\Lcal_i(\beta,\phi_k,\{\hat\eta_{iz}\})\bigr]
    - \log K.
  \label{eq:lse}
\end{equation}
This approximates the exact integral over $\phi_i$ to arbitrary accuracy as
$K \to \infty$.  The required grid size scales as
\begin{equation}
  K \;\gtrsim\; 2\sqrt{\bar N},
  \label{eq:ntheta}
\end{equation}
where $\bar N$ is the mean detected atom count per shot per AI.  In practice
$K = 512$ is used.

\subsection{Profile $\eta_{iz}$}

Each $\eta_{iz}$ has 8 components; for a $32\times32$ image, there are 2048
Poisson measurements per shot per AI.  The likelihood is extremely sharply
peaked in $\eta_{iz}$, so profiling and marginalising are numerically
equivalent.  We profile: for each shot $i$ and AI $z$, find
\begin{equation}
  \hat\eta_{iz}
  = \arg\max_{\eta}\;
    \Bigl[
      \operatorname{logsumexp}_{k}\!\log\Lcal_i^{(z)}(\eta,\phi_k)
      + \log p(\eta)
    \Bigr],
  \label{eq:eta_profile}
\end{equation}
where $\log\Lcal_i^{(z)}$ is the single-AI, single-shot conditional logL
using \emph{both} ports jointly:
\begin{equation}
  \log\Lcal_i^{(z)}(\eta,\phi_k)
  = \sum_{s \in \{g,e\}} \sum_b
    \Bigl[
      n_{izsb} \log \lambda_{izsb}(\phi_k;\eta)
      - \lambda_{izsb}(\phi_k;\eta)
    \Bigr],
\end{equation}
and $\log p(\eta)$ is a Gaussian prior (Section~\ref{sec:prior}).

Crucially, this profiling objective already marginalises over $\phi_i$ via
logsumexp, using the same phase grid as the signal fit.  The resulting
$\hat\eta_{iz}$ is then fixed for all subsequent evaluations of
$\log\Lcal(\beta)$.

\paragraph{Why $\hat\eta$ is approximately independent of $\beta$.}
The cloud parameters $\eta_{iz}$ control the spatial distribution of atoms
across pixels; $\beta$ controls the port fractions via the fringe.  These two
effects are largely orthogonal, so $\hat\eta_{iz}(\beta)$ changes negligibly
as $\beta$ varies.  This justifies computing $\hat\eta_{iz}$ once (at
arbitrary $\beta$, e.g.\ $\beta = 0$) and treating it as fixed during the
outer $\beta$ optimisation.

\subsection{Prior on Cloud Parameters}
\label{sec:prior}

The Gaussian log-prior on $\eta_{iz}$ is
\begin{align}
  \log p(\eta)
  &= -\frac{\mu_{x_0}^2}{2\sigma_{\mu,\mathrm{pos}}^2}
     -\frac{\mu_{y_0}^2}{2\sigma_{\mu,\mathrm{pos}}^2}
     -\frac{\mu_{v_x}^2}{2\sigma_{\mu,\mathrm{vel}}^2}
     -\frac{\mu_{v_y}^2}{2\sigma_{\mu,\mathrm{vel}}^2} \notag\\
  &\quad
     -\frac{(\sigma_x - \bar\sigma_\mathrm{pos})^2}{2\sigma_{\sigma,\mathrm{pos}}^2}
     -\frac{(\sigma_y - \bar\sigma_\mathrm{pos})^2}{2\sigma_{\sigma,\mathrm{pos}}^2}
     -\frac{(\sigma_{v_x} - \bar\sigma_\mathrm{vel})^2}{2\sigma_{\sigma,\mathrm{vel}}^2}
     -\frac{(\sigma_{v_y} - \bar\sigma_\mathrm{vel})^2}{2\sigma_{\sigma,\mathrm{vel}}^2},
  \label{eq:prior}
\end{align}
where the hyperparameters $(\sigma_{\mu,\mathrm{pos}},\sigma_{\mu,\mathrm{vel}},
\bar\sigma_\mathrm{pos},\bar\sigma_\mathrm{vel},\sigma_{\sigma,\mathrm{pos}},
\sigma_{\sigma,\mathrm{vel}})$ are parsed from the dataset name and can be
overridden by command-line arguments.

The optimisation uses a log-sigma parameterisation of the cloud widths:
$\tilde\sigma_x = \log\sigma_x$ etc., so that positivity of widths is enforced
by the domain of $\R$ without explicit bounds.  Bounded L-BFGS-B (rather
than Nelder--Mead) is used, with bounds on the log-widths derived from the
physical constraints of the image field of view.

\paragraph{Initialisation.}
Two strategies are supported:
\begin{itemize}
  \item \emph{Oracle init}: $\hat\eta_{iz}^{(0)} =$ true cloud parameters
        read from the H5 file.  Used for validation; referred to as
        \texttt{--init-from-true}.
  \item \emph{Prior-mean init}: $\hat\eta_{iz}^{(0)} = (0,0,0,0,\bar\sigma_\mathrm{pos},
        \bar\sigma_\mathrm{pos},\bar\sigma_\mathrm{vel},\bar\sigma_\mathrm{vel})$.
\end{itemize}

%%============================================================
\section{Full Marginal Log-Likelihood}
%%============================================================

\subsection{Rotation Trick for AI $z=100$}

For AI $z=100$, the effective phase at grid point $k$ in shot $i$ is
$\phi_k + \delta\phi_i(\beta)$.  Rather than recomputing the ACS for each
$(\delta\phi_i, k)$, we apply a rotation to the precomputed ACS:
\begin{align}
  C^{c,\mathrm{eff}}_{zsb}(\beta)
  &= C^c_{zsb}\cos(\delta\phi_i) + C^s_{zsb}\sin(\delta\phi_i), \\
  C^{s,\mathrm{eff}}_{zsb}(\beta)
  &= -C^c_{zsb}\sin(\delta\phi_i) + C^s_{zsb}\cos(\delta\phi_i).
\end{align}
This follows from the angle-addition identity and means that the per-evaluation
cost for Z100 is a 2D rotation of the precomputed ACS tensors, not a new PSMAP
evaluation.

\subsection{Per-Shot Marginal Log-Likelihood}

Given precomputed ACS $\{A_{zsb}, C^c_{zsb}, C^s_{zsb}\}$ and scale factor
$L_{iz}$ at $\hat\eta_{iz}$, the conditional log-likelihood at phase grid
point $k$ is:
\begin{align}
  \log\Lcal_i(\beta,\phi_k)
  &= \sum_z \sum_{s \in \{g,e\}} \sum_b
     \Bigl[n_{izsb}\log\lambda_{izsb}(\phi_k)
           - \lambda_{izsb}(\phi_k)\Bigr],
  \label{eq:ll_phik}
\end{align}
where for Z0: $\lambda_{igsb}(\phi_k) = L_{i,0}\,[A_{0sb} +
C^c_{0sb}\cos\phi_k + C^s_{0sb}\sin\phi_k]$, and for Z100 the rotated ACS
coefficients are used.  The marginal over $\phi_i$ is then~\eqref{eq:lse}.

\subsection{Full Log-Likelihood}

Since shots are independent given $\beta$:
\begin{equation}
  \log\Lcal(\beta)
  = \sum_{i=1}^{N_\mathrm{shots}} \log\Lcal_i^\mathrm{marg}(\beta).
  \label{eq:full_ll}
\end{equation}

%%============================================================
\section{Implementation}
\label{sec:impl}
%%============================================================

\subsection{ACS Pre-Computation (Once per Shot per AI)}

The following steps are performed once per shot per AI and the results cached
to disk as a compressed NumPy archive (\texttt{.npz}) keyed by
\texttt{run/site/bins/n\_quad/K}:

\begin{enumerate}

  \item \textbf{Cloud nuisance profiling.}
        Solve~\eqref{eq:eta_profile} with L-BFGS-B (up to 200 iterations,
        \texttt{ftol}$=10^{-10}$, \texttt{gtol}$=10^{-8}$, 20 line-search steps).
        The objective evaluates the logsumexp over $K$ phase-grid points for
        both ports jointly at the current $\eta$ via \texttt{SurrogatePixelACS}.

  \item \textbf{ACS at $\hat\eta_{iz}$.}
        Call \texttt{SurrogatePixelACS.pixel\_acs}$(\hat\eta_{iz})$ to obtain
        tensors $\{A_{zsb}, C^c_{zsb}, C^s_{zsb}\}_{s \in \{g,e\}}$ of shape
        $(N_b,)$ each, on GPU.

  \item \textbf{Scale factor.}
        Compute $L_{iz}$ via~\eqref{eq:Lfix}.

  \item \textbf{Cache.}
        Store ACS tensors, $L_{iz}$, $\hat\eta_{iz}$, and raw pixel counts
        $(n_{izgsb}, n_{izesb})$ to \texttt{.npz}.

\end{enumerate}

\subsection{Signal Fit (Per Run)}

Given the cached ACS tables for all shots in a run, fit $\beta = (A_s, A_c)$:

\begin{enumerate}

  \item \textbf{Signal phases.}
        $\delta\phi_i = A_s\sin(2\pi f\,t_i) + A_c\cos(2\pi f\,t_i)$
        for $i = 1,\ldots,N_\mathrm{shots}$.

  \item \textbf{Rotation trick.}
        For each shot $i$, rotate the Z100 ACS tensors by $\delta\phi_i$.

  \item \textbf{Expected counts.}
        For each phase point $\phi_k$, compute $\lambda_{izsb}(\phi_k)$ via
        \eqref{eq:lambda_analytic} as a GPU broadcast:
        \texttt{lam[k, b] = L * max(A[b] + Cc[b]*cos\_phi[k]
        + Cs[b]*sin\_phi[k], 1e-300)}.

  \item \textbf{Conditional logL.}
        Compute~\eqref{eq:ll_phik} for each $(i, k)$ pair.

  \item \textbf{Phase marginalisation.}
        Apply logsumexp over $k$ and subtract $\log K$.

  \item \textbf{Shot sum and optimisation.}
        Sum over shots; negate and pass to L-BFGS-B with bounds
        $A_s \in [0, \pi]$, $A_c \in [-\pi, \pi]$.

\end{enumerate}

\subsection{Multi-Start Strategy for $\beta$}

The marginal log-likelihood $\log\Lcal(A_s, A_c)$ is smooth but can be flat
near the null $(0,0)$ when the signal is weak.  We use multi-start L-BFGS-B
from amplitude starting values $A_s^{(0)} \in \{0, 0.03, 0.07, 0.10, 0.15\}$
at $A_c^{(0)} = 0$, take the best solution, and report convergence diagnostics.

\subsection{Computational Cost}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
Step & Phase & Typical wall-time \\
\midrule
\texttt{SurrogatePixelACS} construction   & Once (shared) & $\sim\!12$\,s \\
Cloud profiling ($K{=}512$ logsumexp)     & Per shot/AI   & $\sim\!0.5$--$2$\,s \\
All shots profiling + ACS cache           & Per run       & $\sim\!1$--$2$\,min \\
Signal fit (50 shots, multi-start)        & Per run       & $\sim\!40$\,s \\
\bottomrule
\end{tabular}
\caption{Typical timings on one GPU (A100/V100 class) with $K=512$, $N_b=1024$
($32\times32$ bins), $n_\mathrm{quad}=20{,}000$, $N_\mathrm{shots}=50$.
Cloud profiling timings dominate on the first pass; subsequent runs load from
the \texttt{.npz} cache in milliseconds.}
\label{tab:cost}
\end{table}

%%============================================================
\section{Maximum Likelihood Estimation}
%%============================================================

\subsection{Science Parameters}

With $f$ fixed, the only free parameters are $\beta = (A_s, A_c)$.
The MLE is
\begin{equation}
  \hat\beta = \arg\max_\beta\; \log\Lcal(\beta),
  \label{eq:mle}
\end{equation}
solved by multi-start L-BFGS-B (Section~\ref{sec:impl}).  The recovered
amplitude and phase are $\hat A = \sqrt{\hat A_s^2 + \hat A_c^2}$ and
$\hat\varphi = \arctan2(\hat A_c, \hat A_s)$.

\subsection{True Signal Parameterisation}

The simulated signal is generated as
$\delta\phi_i = A_\mathrm{true}\sin(2\pi f t_i + \varphi_\mathrm{true})$.
Expanding via angle addition:
\begin{equation}
  A_s^\mathrm{true} = A_\mathrm{true}\cos\varphi_\mathrm{true},
  \qquad
  A_c^\mathrm{true} = A_\mathrm{true}\sin\varphi_\mathrm{true},
\end{equation}
so the ground-truth $(A_s, A_c)$ is read directly from the H5 file attributes
\texttt{signal\_amp} and \texttt{signal\_phase}.  Note that $\delta\phi_i$ is
stored only in the Z100 H5 file (it is zero in Z0 by construction).

%%============================================================
\section{Validation}
%%============================================================

Running \texttt{fit\_surrogate\_profiled\_signal.py} with oracle cloud
initialisation (\texttt{--init-from-true}) on the default 1\,M-atom dataset
($N_\mathrm{shots}=50$, $K=512$, $n_\mathrm{quad}=20{,}000$) gives:

\begin{itemize}
  \item Signal amplitude RMSE $\approx 0.0013$\,rad
        (1.3\,\% of true $A=0.1$\,rad).
  \item Signal phase RMSE $\approx 0.009$\,rad.
  \item Cloud COM RMSE: $\sim\!2$--$5$\,$\mu$m for position,
        $\sim\!1$--$4$\,$\mu$m/s for velocity.
  \item Cloud width RMSE: $\sim\!0.1$--$0.3$\,$\mu$m.
  \item L-BFGS-B converges in $\sim\!5$ iterations and $\sim\!25$ function
        evaluations for both the cloud profiling and the signal fit.
\end{itemize}

\end{document}
